---
title: "HW4"
author: "STAT 28"
output:
  pdf_document
header-includes:
- \usepackage{framed}
- \usepackage{xcolor}
- \let\oldquote=\quote
- \let\endoldquote=\endquote
- \colorlet{shadecolor}{orange!15}
- \renewenvironment{quote}{\begin{shaded*}\begin{oldquote}}{\end{oldquote}\end{shaded*}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,tidy.opts=list(width.cutoff=50))
```

We are going to use the fitbit data again. However, I have changed some of the column names, so make sure you use the dataset attached with this homework. 

Below is code to read in the data. The rest of the code will convert the `Date` variable into a standard date class used by R, and then create variables that give the day of the week and the month of the date. 

```{r readinFitbit}
fitbit<-read.csv("fitbit.csv",header=TRUE)
fitbit$Date<-as.Date(as.character(fitbit$Date),format="%d-%m-%Y")
fitbit$Day<-factor(weekdays(fitbit$Date),levels=weekdays(x=as.Date(seq(7), origin="1950-01-01")))
fitbit$Month<-factor(months(fitbit$Date),levels=month.name)
```

Notice how I explicitly give the function `factor` the levels to expect. This allows me to define what order they will be in, so I can force them to be in a proper order for days/months (to save myself typing and possible typos, I used built in functions in R to find the names of months and weeks in the right order, but I could have just typed them out too). 


**Question 1:** (5 points) The below code changes the dataset to convert the minutes to percentages of the total, like the last homework, in addition to other changes. Describe what lines 3-5 do. You may need to look at the `help` of the function `gsub` and `abbreviate`. 

```{r modifyDataSet}
totalNumberOfMinutes <- fitbit$MinutesOfSleep + fitbit$MinutesOfLightActivity + fitbit$MinutesOfModerateActivity + fitbit$MinutesOfSedentaryActivities + fitbit$MinutesOfIntenseActivity #LINE 1
absoluteMinNames<-c("MinutesOfSedentaryActivities","MinutesOfLightActivity","MinutesOfModerateActivity" , "MinutesOfIntenseActivity" , "activityCalories","MinutesOfSleep", "MinutesOfBeingAwake" ,        
"NumberOfAwakings","MinutesOfRest") #LINE 2
fitbit[,absoluteMinNames]<-fitbit[,absoluteMinNames]/totalNumberOfMinutes  #LINE 3
names(fitbit)<-gsub("Minutes","pctMin",names(fitbit)) #LINE 4
names(fitbit)<-abbreviate(names(fitbit),10) #LINE 5
```

> My answer: Line 4 replaces any of the text in the fitbit names where "Minute" appears and replaces it with "pctMin".  Line 5 abbreviates all the fitbit names to be 10 characters or less by first eliminating spaces, then lower case vowels, then lower case consonants.

In future questions make sure you use this modified data.frame.

**Question 2:** (10 points) Create a pairs plot using the varaibles `caloriesBurned`,  `steps`,  `pctMinOfIntenseActivity`,  `Month` and `Day`. Instead of the default, put boxplots on both the upper and lower diagonal for the categorical variables (not just the lower). Comment which variables appear to have a strong pairwise relationship and which do not. 

```{r pairsPlot}
#Code for pairs plot here.
variables<-c("calorsBrnd","steps","pctMnOfInA","Month","Day")
shortened<-fitbit[,variables]
library(gpairs)
gpairs(shortened,
       upper.pars = list(
                        conditional = "boxplot"
                         ),
       lower.pars = list(
                         conditional = "boxplot"),
       diagonal = "default")
```

> My answer:  Calories burned and steps have a strong connection to one another but not for percent minutes of intense activity and steps.

**Question 3:** (10 points) Use `corrgram` to plot all of the continuous variables in the dataset [Hint: look carefully at the variables, including `distance`, before you start plotting ]. Comment on what you see. 

```{r corrPlot}
#Code for corrgram plot here.
continuous<-fitbit[,c(2:3,5:14)]
corrgram(continuous)
```

> My answer: As expected there is a strong negative associaiton between minutes of sedentary activity with the other levels of activity and there is a strong positive association between calories burned and steps taken.  There are negative associations between the different levels of activity and the time spent asleep and the time spent resting.

**Question 4:** 

(a) (5 points) Plot the percent of sedentary activity as a function of `Date`, making sure to remove the (practically) zero values as in HW3 [See solutions of HW3 if you were not able to do this successfully yourself]. Color the points according the day of the week and give a legend. We have defined the colors for you in the following chunk with the vector `colWeek`.

```{r plotByDate}
colWeek<-palette()[1:nlevels(fitbit$Day)]
names(colWeek)<-levels(fitbit$Day)
colWeek
# code to plot sedentary against date
nonzero<-fitbit[fitbit$pctMnOfSdA>0.2,]

#Creates a variable that attaches the colors to the days

nonzero$color<-ifelse(nonzero$Day=="Monday","black",ifelse(nonzero$Day=="Tuesday","red",ifelse(nonzero$Day=="Wednesday","green3",ifelse(nonzero$Day=="Thursday","blue",ifelse(nonzero$Day=="Friday","cyan",ifelse(nonzero$Day=="Saturday","magenta",ifelse(nonzero$Day=="Sunday","yellow","")))))))

plot(nonzero$Date,nonzero$pctMnOfSdA,col=nonzero$color)

legend("topleft",pch=c(1,1),legend=c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),col=c(colWeek),cex = 0.3)

```

(b) (10 points) Draw loess lines for for each day of the week that fit the percent of sedentary activity as a function of date. The loess lines should all be on the same plot.  Color each line by the day of the week, as in the previous plot, and include a legend. Use the `by` function to do this. 

The first line of your code should set up a blank plot. Do this by reusing your code from (a) above (i.e. that plots sedentary against the data), but include the argument `type="n"` in your plot command. This instructs R to set up the axes, etc. but not actually plot the points. 

```{r plotLoessPerDay}
# code to plot loess lines

colWeek<-palette()[1:nlevels(nonzero$Day)]
names(colWeek)<-levels(nonzero$Day)

plot(nonzero$Date,nonzero$pctMnOfSdA,col=nonzero$color,ylab="PctMnOfSedAct",xlab="Date",type = "n")
by(nonzero,nonzero$Day,function(x){
  lines(loess.smooth(x=nonzero$Date, y=nonzero$pctMnOfSdA),col=colWeek[unique(nonzero$Day)])
})                                                    

```

(c) (5 points) Comment on whether there are any difference due to the day of the week.

> My answer